---
title: "Milestone 4"
author: "Team 20"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(knitr)
library(tidyverse)
library(readr)
library(kableExtra)
library(ggpubr)
library(ggplot2)

#debug LaTeX
#tinytex::tlmgr_update()
#options(tinytex.verbose = TRUE)
```


```{r, echo = FALSE,results="hide",include=FALSE}
#air quality 
air_qual <- read_csv("air_qual.csv",
                    col_names = TRUE, col_types = NULL, na = c("", "NA"))
ca_air_qual <- rename(air_qual, county = "California County", 
                      zip = "ZIP", 
                      avg_PM2.5_per_zip = "Avg PM2.5 Per ZIP")
ca_air_qual[,3] <- round(ca_air_qual$avg_PM2.5_per_zip, 3)

#mort data 
mort_by_zip <- read_csv("mort_by_zip.csv",
  na = c("", "NA"))
mort_by_zip <- select(mort_by_zip, "Year", "ZIP_Code", "Cause", "Cause_Desc", "Count")
mort_by_year <- filter(mort_by_zip, between(Year, 2009, 2018) )
mort_by_cause <- filter(mort_by_year, Cause == "CLD")
head(mort_by_cause)
dim(mort_by_cause)

```

```{r, echo = FALSE,results="hide",include=FALSE}
#grouping all zipcodes 
air_qual_grouped_zips <- ca_air_qual %>% 
  group_by(zip,county) %>% 
  summarise(avg_PM_per_zip = mean(avg_PM2.5_per_zip))

```


```{r, echo = FALSE,results="hide",include=FALSE}
air_qual_grouped_zips <- air_qual_grouped_zips %>% 
  mutate(critical_level_pollution =  case_when(avg_PM_per_zip >= 11.84~1))

air_qual_grouped_zips$critical_level_pollution[is.na(air_qual_grouped_zips$critical_level_pollution)] <- 0

#creating a subset of zip codes that have PM2.5 above critical value 
air_qual_by_75_percent <- air_qual_grouped_zips %>%
  filter(critical_level_pollution == 1)

```


```{r, echo = FALSE,results="hide",include=FALSE}
#remove duplicates of zip code
length(duplicated(air_qual_by_75_percent$zip))
length(unique(air_qual_by_75_percent$zip))
air_qual_by_75_percent<-air_qual_by_75_percent[!duplicated(air_qual_by_75_percent$zip), ]

```

```{r, echo = FALSE,results="hide",include=FALSE}
mort_grouped_zip <- mort_by_cause %>% 
  group_by(ZIP_Code, Cause) %>% 
  summarise(sum_count = sum(Count))

head(mort_grouped_zip)
dim(mort_grouped_zip)
```


```{r, echo = FALSE,results="hide",include=FALSE}
mort_matched_zips <- mort_grouped_zip %>% 
  filter(ZIP_Code %in% air_qual_by_75_percent$zip) 

mort_matched_zips 
```


```{r, echo = FALSE,results="hide",include=FALSE}
final_data <- cbind(air_qual_by_75_percent, mort_matched_zips)
final_data <- final_data %>% 
  select( county, zip, avg_PM_per_zip, sum_count)

#cleaning 
final_data$sum_count[is.na(final_data$sum_count)] <- NA
final_data$avg_PM_per_zip <- round(final_data$avg_PM_per_zip, 2)
final_data$zip <-as.character(final_data$zip)
head(final_data)
#final_data[is.na(final_data)] <- 0 

```

# TABLES

## Table 1 and 2 display average pollution measured by Particulate Matter 2.5 (PM2.5) and deaths due to Chonric Lower Respiratory Disease (CLRD) by zip code and county levels in California.


### Table 1: Average pollution and CLRD deaths per zip code

```{r}
avgpm_mort_zip2<-final_data %>% 
  group_by(zip) %>% 
  summarise(county, average_PM2.5=mean(avg_PM_per_zip), death = sum(sum_count, na.rm = T)) %>%
  filter(death > 0)

avgpm_mort_zip2 <- avgpm_mort_zip2 %>% arrange(desc("average_PM2.5"))

kable(avgpm_mort_zip2, booktabs =T, 
      col.names=c("Zip Code","County", "Average PM2.5","CLDRM Counts"),
      align='lccc',
      caption="Table 1: Chronic Lower Respiratory Disease (CLRD) Mortality by California 
      zip codes with critically high levels of Particulate Matter 2.5 (PM2.5),
      2009-2018"
      ) %>%
  kable_styling(full_width=F) %>%
  kable_styling(position="left") %>%
  kable_styling(font_size=10) %>%
  column_spec(1, bold=TRUE) %>%
  column_spec(4, color=if_else(avgpm_mort_zip2$death>278,"red","black")) %>%
  footnote(
    general="Counties with the most number of zip codes with both critical levels
    of PM2.5 and CLRD deaths include Los Angeles (13), San Bernadino (10), and 
    Kern (7) counties. On the other hand, zip codes that hold the highest death 
    counts in the 75th percentile belong to many counties: Kern, Los Angeles, 
    Stanislaus, Tulare, Riverside, San Diego, and Kings counties, with one Kern 
    zip code (93308) holding the highest death count of 469. Kern zip codes also 
    hold the top 7 highest PM2.5 levels ranging between 18.76-19.31.",
    general_title="Interpretation",
    title_format=c("italic"),
    threeparttable=T) %>%
   footnote(
    general="CLRD mortality counts highlighted in red represent death counts in the 
    75th percentile. Only zip codes with critical levels of PM2.5 are included in 
    this table, and zip codes with no reported deaths between 2009-2018 were excluded.",
    general_title="Legend",
    title_format=c("italic"),
    threeparttable=T)
 

```

\newpage

### Table 2: Average pollution and CLRD deaths by county

```{r}
avgpm_mort_county<-final_data %>% 
  group_by(county) %>% 
  summarise(average_PM2.5= round(mean(avg_PM_per_zip),digits = 2), 
            death = sum(sum_count, na.rm = T))

avgpm_mort_county <- avgpm_mort_county %>% arrange(desc(death))

kable(avgpm_mort_county, booktabs =T, 
      col.names=c("County", "Average PM2.5","CLRD Mortality Counts"),
      align='lcc',
      caption="Table 2: Chronic Lower Respiratory Disease (CLRD) Mortality by 
      California counties with critically high levels of Particulate Matter 2.5 
      (PM2.5), 2009-2018") %>%
  kable_styling(full_width=F) %>%
  kable_styling(position="left") %>%
  kable_styling(font_size=10) %>%
  column_spec(1, bold=TRUE) %>%
  footnote(
    general="In accordance with the table above, Los Angeles, San Bernardino, and
    Kern counties hold the highest death counts. Counties with the highest average
    PM2.5 levels are Kern, Kings, Tulare, and Fresno (17.57, 15.86, 15.56, and D
    15.24, respectively). Despite having critical levels of PM2.5, the counties 
    of Imperial, Madera, and Orange had no reported CLRD deaths between 2009-2018.",
    general_title="Interpretation",
    title_format=c("italic"),
    threeparttable=T) %>%
   footnote(
    general="Zip code level PM2.5 and CLRD death data are averaged by county. 
    Only counties with critical levels of PM2.5 are included in this table, and 
    zip codes with no reported deaths between 2009-2018 were excluded.",
    general_title="Legend",
    title_format=c("italic"),
    threeparttable=T)
```

\newpage

### CHARTS

## The following two charts seek to explore the relationship between the continuous PM2.5 values and CLRD Mortality counts occurring in all the identified critical category zip codes 


```{r}
final_data_1<-na.omit(final_data)

final_data_1$zip<-as.factor(final_data_1$zip)
  ggplot(final_data_1)  + 
  geom_bar(aes(x = reorder(zip, -sum_count), y=sum_count,color = "sum_count"),stat="identity", fill="cyan")+ 
  geom_line(aes(x = reorder(zip, -sum_count),y=avg_PM_per_zip*10),stat="identity",color="red",
            group=1,ylim=c(0,20))+
  labs(title= "Figure 1: Chronic Lower Respiratory Disease (CLRD) mortality 
       and Particulate Matter 2.5 (PM2.5) air pollution measurements per zip code 
       among those with critical levels of 2.5PM (California 2009-2018)",
       x="Zip Code",y="CLRD Mortality",
        caption = 
         "Figure 1 shows the CLRD mortality and PM2.5 air pollution measurements in zip 
         codes where PM2.5 measurements are in the critical category (the highest quantile). 
         The blue bars represent CLRD death counts while the rea line represents PM2.5 air 
         pollution measurements this plot we cannot see a strong correlation between 
         CLRD mortality and PM2.5 measurements." )+
   scale_y_continuous(sec.axis=sec_axis(~.*0.1,name="PM2.5 Measurements"))+
    theme(axis.text.x=element_text(angle=90,colour="grey20",face="bold",size=8))  + theme(plot.caption = element_text(hjust = 0))+ theme(legend.position="bottom")+
   scale_color_discrete(label="Chronic Lower Respiratory Disease (CLRD) mortality")
  
```


\newpage

```{r}
ggplot(data = final_data, aes(x = avg_PM_per_zip, y = sum_count)) +
  geom_point(na.rm = T) + 
  geom_smooth(method = "lm", na.rm = T, se = T, level = .95) + theme_bw() + 
  labs(title = "Figure 2: Total Chronic Lower Respiratory Disease (CLRD) mortality by 
               Particulate Matter 2.5 (PM2.5) air pollution per zip code among those with 
               critical levels of PM2.5 (California 2009-2018) ",
       x= "PM2.5 pollution (Critical levels)", y= "Total CLRD mortality count ", 
       caption = "Figure 2 shows a positive association between higher levels of PM2.5 air pollution
       and increased levels of CLRD mortality per zip code. Our correlation coefficient value is 0.27, which
       indicates a small correlation but positive correlation between critical levels of PM2.5 air pollution
       and CLDR mortality. The p-value for our correlation is 0.039, meaning that although our correlation is 
       small, it is still statistically significant.")  +
  stat_cor(method = "pearson",na.rm = T) + theme(plot.caption = element_text(hjust = 0))

```



