---
title: "Milestone 6"
author: "Team 20"
date: 12/13/2021
output:
  html_document:
    df_print: paged

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(knitr)
library(tidyverse)
library(readr)
library(kableExtra)
library(ggpubr)
library(ggplot2)

```


```{r, echo = FALSE,results="hide",include=FALSE}
#air quality 
air_qual <- read_csv("air_qual.csv",
                    col_names = TRUE, col_types = NULL, na = c("", "NA"))
ca_air_qual <- rename(air_qual, county = "California County", 
                      zip = "ZIP", 
                      avg_PM2.5_per_zip = "Avg PM2.5 Per ZIP")
ca_air_qual[,3] <- round(ca_air_qual$avg_PM2.5_per_zip, 3)

#mort data 
mort_by_zip <- read_csv("mort_by_zip.csv",
  na = c("", "NA"))
mort_by_zip <- select(mort_by_zip, "Year", "ZIP_Code", "Cause", "Cause_Desc", "Count")
mort_by_year <- filter(mort_by_zip, between(Year, 2009, 2018) )
mort_by_cause <- filter(mort_by_year, Cause == "CLD")
head(mort_by_cause)
dim(mort_by_cause)

```

```{r, echo = FALSE,results="hide",include=FALSE}
#grouping all zipcodes 
air_qual_grouped_zips <- ca_air_qual %>% 
  group_by(zip,county) %>% 
  summarise(avg_PM_per_zip = mean(avg_PM2.5_per_zip))

```


```{r, echo = FALSE,results="hide",include=FALSE}
air_qual_grouped_zips <- air_qual_grouped_zips %>% 
  mutate(critical_level_pollution =  case_when(avg_PM_per_zip >= 11.84~1))

air_qual_grouped_zips$critical_level_pollution[is.na(air_qual_grouped_zips$critical_level_pollution)] <- 0

#creating a subset of zip codes that have PM2.5 above critical value 
air_qual_by_75_percent <- air_qual_grouped_zips %>%
  filter(critical_level_pollution == 1)

```


```{r, echo = FALSE,results="hide",include=FALSE}
#remove duplicates of zip code
length(duplicated(air_qual_by_75_percent$zip))
length(unique(air_qual_by_75_percent$zip))
air_qual_by_75_percent<-air_qual_by_75_percent[!duplicated(air_qual_by_75_percent$zip), ]

```

```{r, echo = FALSE,results="hide",include=FALSE}
mort_grouped_zip <- mort_by_cause %>% 
  group_by(ZIP_Code, Cause) %>% 
  summarise(sum_count = sum(Count))

head(mort_grouped_zip)
dim(mort_grouped_zip)
```


```{r, echo = FALSE,results="hide",include=FALSE}
mort_matched_zips <- mort_grouped_zip %>% 
  filter(ZIP_Code %in% air_qual_by_75_percent$zip) 

mort_matched_zips 
```


```{r, echo = FALSE,results="hide",include=FALSE}
final_data <- cbind(air_qual_by_75_percent, mort_matched_zips)
final_data <- final_data %>% 
  select( county, zip, avg_PM_per_zip, sum_count)

#cleaning 
final_data$sum_count[is.na(final_data$sum_count)] <- NA
final_data$avg_PM_per_zip <- round(final_data$avg_PM_per_zip, 2)
final_data$zip <-as.character(final_data$zip)
head(final_data)
#final_data[is.na(final_data)] <- 0 

```

\newpage

# Problem Statement

There are many known health issues associated with air pollution, often measured by Particulate Matter (PM), a term for solid or liquid particles or droplets that are found in the air (such as dust, smog, or smoke). PM of size 2.5 micrometers or smaller are very fine inhalable particles that are particularly damaging to lung health*. Our team seeks to investigate the potential links between Air Quality Levels and Chronic Lower Respiratory Disease (CLRD) Mortality  in California. 
 
Currently, PM2.5 levels of under 12.0 are considered “good,” and under 35.4 are considered “moderate” according to the US Environmental Protection Agency. It is not until we reach levels over 35.5 that air quality is considered “unhealthy for sensitive groups,” and over 55.5 to be considered “unhealthy.” Our team believes that this threshold of “safe” PM2.5 levels should be drastically lowered in California. We will investigate aggregate statewide PM2.5 measurements and CLRD Mortality counts 1) at a zip code level to determine where further research should be conducted, and 2) examine a potential correlation between zip codes with critically high PM2.5 levels and CLRD deaths from 2009-2018.

* Environmental Protection Agency. (n.d.). Particulate Matter (PM) Pollution. EPA. Retrieved November 21, 2021, from https://www.epa.gov/pm-pollution/particulate-matter-pm-basics.

\newpage

# Methods

### Data Sources
Exposure data: This dataset comes from CalEnviroScreen3.0 dataset from the California Environmental Protection Agency (CalEPA). In January 2017, the Office of Environmental Health Hazard Assessment (OEHHA), on behalf of the California Environmental Protection Agency (CalEPA) released this dataset.  It contains data from  three columns, county, zip code and average PM2.5 per zip code. 
Outcome data: This dataset contains counts of deaths for California residents by ZIP Code based on information entered on death certificates from 2009-2018.

### Cleaning and Creating New Variable
We first removed the NA from our exposure data and obtained the 75th percentile (11.83) of the average PM2.5 per zip code. We then grouped our exposure data by zip code and created a new binary column to indicate whether or not the zip has PM2.5 measurement in the critical category of PM2.5 (the 75th percentile). We created a subset of our data which only included zip codes in the critical category. Duplicated zip codes were eliminated from the dataset.

We grouped our outcome data by zip code and created a new column called total count containing the total number of CLRD deaths in each zip code from 2009 to 2018. We then filtered zip codes so that only those that are both in our exposure dataset and outcome dataset were retained. Zip codes that were not in the exposure dataset were not included in this new subset.

We combined our exposure data, zip codes and corresponding PM2.5 measurements in the critical category, and the outcome data, zip codes and corresponding CLRD deaths, into one final dataset. Columns of county, zip code, PM2.5 measurements and total number of CLRD deaths were retained. NA were removed from this final dataset. Zip codes were stored as factors. PM2.5 measurements were rounded to two decimal places.

### Analytic Methods
A plot containing a bar chart and a line chart was created from our final dataset. The bar chart represents the total number of CLRD deaths in each zip code while the line chart represents the PM2.5 measurements in each zip code. Then linear regression was used to test the relationship between PM2.5 pollution levels and total Chronic Lower Respiratory Disease Mortality among zip codes in the critical category of PM2.5. 

\newpage

# Results
Table 1 provides a list of all of the zip codes with critical values of PM2.5 in California, their air pollution level, and the number of deaths caused by Chronic Lower Respiratory Disease(CLRD) in these regions between the years 2009-2018. Table 2 collapses this information to demonstrate average PM2.5 and total CLRD mortality on the county level. Table 1 and Table 2 facilitate the identification of zip codes with the highest levels of PM2.5 pollution or the largest CLRD mortality counts. These tables can be posted on our nonprofit’s website to help inform our audience on this issue. Because further research on all of these zip codes would be costly, we recommend a targeted investigation of the zip codes that ranked among the top 25 most burdened for both PM2.5 pollution and Chronic Lower Respiratory Disease Mortality. These zip codes are "93307" (Kern),  "93306" (Kern), "93304" (Kern), "93308" (Kern), "93309" (Kern),  "93277" (Tulare),  "93274" (Tulare),  and "93230" (Kings).

```{r, echo = FALSE,results="hide",include=FALSE}
avgpm_mort_zip2<-final_data %>% 
  group_by(zip) %>% 
  summarise(county, average_PM2.5=mean(avg_PM_per_zip), death = sum(sum_count, na.rm = T)) %>%
  filter(death > 0)

avgpm_mort_zip2 <- avgpm_mort_zip2 %>% arrange(desc("average_PM2.5"))

table1 <- kable(avgpm_mort_zip2, booktabs =T, 
      col.names=c("Zip Code","County", "Average PM2.5","CLDRM Counts"),
      align='lccc',
      caption="Table 1: Chronic Lower Respiratory Disease (CLRD) Mortality by California 
      zip codes with critically high levels of Particulate Matter 2.5 (PM2.5),
      2009-2018"
      ) %>%
  kable_styling(full_width=F) %>%
  kable_styling(position="left") %>%
  kable_styling(font_size=10) %>%
  column_spec(1, bold=TRUE) %>%
  column_spec(4, color=if_else(avgpm_mort_zip2$death>278,"red","black")) %>%
  footnote(
    general="Counties with the most number of zip codes with both critical levels
    of PM2.5 and CLRD deaths include Los Angeles (13), San Bernadino (10), and 
    Kern (7) counties. On the other hand, zip codes that hold the highest death 
    counts in the 75th percentile belong to many counties: Kern, Los Angeles, 
    Stanislaus, Tulare, Riverside, San Diego, and Kings counties, with one Kern 
    zip code (93308) holding the highest death count of 469. Kern zip codes also 
    hold the top 7 highest PM2.5 levels ranging between 18.76-19.31.",
    general_title="Interpretation",
    title_format=c("italic"),
    threeparttable=T) %>%
   footnote(
    general="CLRD mortality counts highlighted in red represent death counts in the 
    75th percentile. Only zip codes with critical levels of PM2.5 are included in 
    this table, and zip codes with no reported deaths between 2009-2018 were excluded.",
    general_title="Legend",
    title_format=c("italic"),
    threeparttable=T)
 

```

```{r}
table1
```

\newpage

 

```{r, echo = FALSE,results="hide",include=FALSE}
avgpm_mort_county<-final_data %>% 
  group_by(county) %>% 
  summarise(average_PM2.5= round(mean(avg_PM_per_zip),digits = 2), 
            death = sum(sum_count, na.rm = T))

avgpm_mort_county <- avgpm_mort_county %>% arrange(desc(death))

table2 <- kable(avgpm_mort_county, booktabs =T, 
      col.names=c("County", "Average PM2.5","CLRD Mortality Counts"),
      align='lcc',
      caption="Table 2: Chronic Lower Respiratory Disease (CLRD) Mortality by 
      California counties with critically high levels of Particulate Matter 2.5 
      (PM2.5), 2009-2018") %>%
  kable_styling(full_width=F) %>%
  kable_styling(position="left") %>%
  kable_styling(font_size=10) %>%
  column_spec(1, bold=TRUE) %>%
  footnote(
    general="In accordance with the table above, Los Angeles, San Bernardino, and
    Kern counties hold the highest death counts. Counties with the highest average
    PM2.5 levels are Kern, Kings, Tulare, and Fresno (17.57, 15.86, 15.56, and D
    15.24, respectively). Despite having critical levels of PM2.5, the counties 
    of Imperial, Madera, and Orange had no reported CLRD deaths between 2009-2018.",
    general_title="Interpretation",
    title_format=c("italic"),
    threeparttable=T) %>%
   footnote(
    general="Zip code level PM2.5 and CLRD death data are averaged by county. 
    Only counties with critical levels of PM2.5 are included in this table, and 
    zip codes with no reported deaths between 2009-2018 were excluded.",
    general_title="Legend",
    title_format=c("italic"),
    threeparttable=T)
```

```{r}
table2
```


\newpage

Figure 1 visualizes average PM2.5 levels and CLRD mortality among zip codes with a critical value of PM2.5 pollution. To identify the extent of association between the two, we used linear regression to measure a correlation value, as seen in Figure 2. Our intuition was that we would see CLDR mortality increase as levels of air pollution increase. Our linear regression model found a correlation value of 0.27, which indicates a small association but statistically significant correlation (p < 0.5).

```{r, echo = FALSE,results="hide",include=FALSE}
final_data_1<-na.omit(final_data)

final_data_1$zip<-as.factor(final_data_1$zip)

figure1 <- ggplot(final_data_1)  + 
  geom_bar(aes(x = reorder(zip, -sum_count), y=sum_count,color = "sum_count"),stat="identity", fill="cyan")+ 
  geom_line(aes(x = reorder(zip, -sum_count),y=avg_PM_per_zip*10),stat="identity",color="red",
            group=1,ylim=c(0,20))+
  labs(title= "Figure 1: Chronic Lower Respiratory Disease (CLRD) mortality 
       and Particulate Matter 2.5 (PM2.5) air pollution measurements per zip code 
       among those with critical levels of 2.5PM (California 2009-2018)",
       x="Zip Code",y="CLRD Mortality",
        caption = 
         "Figure 1 shows the CLRD mortality and PM2.5 air pollution measurements in zip 
         codes where PM2.5 measurements are in the critical category (the highest quantile). 
         The blue bars represent CLRD death counts while the rea line represents PM2.5 air 
         pollution measurements this plot we cannot see a strong correlation between 
         CLRD mortality and PM2.5 measurements." )+
   scale_y_continuous(sec.axis=sec_axis(~.*0.1,name="PM2.5 Measurements"))+
    theme(axis.text.x=element_text(angle=90,colour="grey20",face="bold",size=8))  + theme(plot.caption = element_text(hjust = 0))+ theme(legend.position="bottom")+
   scale_color_discrete(label="Chronic Lower Respiratory Disease (CLRD) mortality")
  
```

```{r}
figure1

```


\newpage

```{r, echo = FALSE,results="hide",include=FALSE}
figure2 <- ggplot(data = final_data, aes(x = avg_PM_per_zip, y = sum_count)) +
  geom_point(na.rm = T) + 
  geom_smooth(method = "lm", na.rm = T, se = T, level = .95) + theme_bw() + 
  labs(title = "Figure 2: Total Chronic Lower Respiratory Disease (CLRD) mortality by 
               Particulate Matter 2.5 (PM2.5) air pollution per zip code among those with 
               critical levels of PM2.5 (California 2009-2018) ",
       x= "PM2.5 pollution (Critical levels)", y= "Total CLRD mortality count ", 
       caption = "Figure 2 shows a positive association between higher levels of PM2.5 air pollution
       and increased levels of CLRD mortality per zip code. Our correlation coefficient value is 0.27, which
       indicates a small correlation but positive correlation between critical levels of PM2.5 air pollution
       and CLDR mortality. The p-value for our correlation is 0.039, meaning that although our correlation is 
       small, it is still statistically significant.")  +
  stat_cor(method = "pearson",na.rm = T) + theme(plot.caption = element_text(hjust = 0))

```

```{r}
figure2
```
\newpage

# Discussion
According to the US Environmental Protection Agency, levels of PM2.5 between 12.1 and 35.4 are considered “moderate”. While the EPA states that these levels of air pollution can be classified as “safe”, we argue that any level of PM2.5 above 11.84 should be reclassified as “unhealthy” in the state of California. Our group believes that truly “safe” levels of PM2.5 are those which demonstrate zero correlation between air pollution and CLRD mortality. However, our analysis suggests that there is a statistically significant correlation between “moderate” levels of air pollution and CLRD mortality (Figure 2). This is further supported by Figure 1, which demonstrates elevated levels of CLDR mortality for all levels of PM2.5 pollution greater than 11.84. We believe further research in the zip codes most burdened by both PM2.5 and CLDR mortality will provide insight into this relationship. We believe this research will also give us information on other potential safety measures California may consider implementing, such as a minimum residential distance from high emission structures. 

