---
title: "Milestone 3: Clean dataset with descriptive statistics for relevant data elements"
author: "Team 20"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(knitr)
library(tidyverse)
library(readr)
library(tableone)
library(kableExtra)
```


#1. Subset rows & columns
```{r }
#air quality 
air_qual<- read_csv("air_qual.csv",
                    col_names = TRUE, col_types = NULL, na = c("", "NA"))
ca_air_qual <- rename(air_qual, county = "California County", 
                      zip = "ZIP", 
                      avg_PM2.5_per_zip = "Avg PM2.5 Per ZIP")
ca_air_qual[,3] <- round(ca_air_qual$avg_PM2.5_per_zip, 3)
head(ca_air_qual)
dim(ca_air_qual)

#mort data 
mort_by_zip <- read_csv("mort_by_zip.csv",na = c("", "NA"))
mort_by_zip <- select(mort_by_zip, "Year", "ZIP_Code", "Cause", "Cause_Desc", "Count")
mort_by_year <- filter(mort_by_zip, between(Year, 2009, 2018) )
mort_by_cause <- filter(mort_by_year, Cause == "CLD")
head(mort_by_cause)
dim(mort_by_cause)

```

\newpage

#2. Creating 2+ new variables (and subsetting by pollution level)

Identify and create a "critical" category of PM2.5 measurements (the highest quantile), and subset those zipcode observations. 

```{r }
#grouping all zipcodes 
air_qual_grouped_zips <- ca_air_qual %>% 
  group_by(zip,county) %>% 
  summarise(avg_PM_per_zip = mean(avg_PM2.5_per_zip))

```



```{r }
quantile(air_qual_grouped_zips$avg_PM_per_zip, c(.10,.25,.5,.75,.90), 
         na.rm = TRUE)
#75th percentile is 11.84
#90th percentile is 12.89

#creating new binary column to indicate whether or not the zip has a critical 
#value of PM2.5 pollution

air_qual_grouped_zips <- air_qual_grouped_zips %>% 
  mutate(critical_level_pollution =  case_when(avg_PM_per_zip >= 11.84~1))

air_qual_grouped_zips$critical_level_pollution[is.na(air_qual_grouped_zips$critical_level_pollution)] <- 0

#creating a subset of zip codes that have PM2.5 above critical value 
air_qual_by_75_percent <- air_qual_grouped_zips %>%
  filter(critical_level_pollution == 1)

```


```{r }
#remove duplicates of zip code
duplicated(air_qual_by_75_percent$zip)
length(duplicated(air_qual_by_75_percent$zip))
length(unique(air_qual_by_75_percent$zip))
air_qual_by_75_percent<-air_qual_by_75_percent[!duplicated(air_qual_by_75_percent$zip), ]

```

Creating a *total count* of CLDRM for the 2009-2018 period.
```{r }
mort_grouped_zip <- mort_by_cause %>% 
  group_by(ZIP_Code, Cause) %>% 
  summarise(sum_count = sum(Count))

head(mort_grouped_zip)
dim(mort_grouped_zip)
```

subset of mortality data to match pollution data 
```{r }
mort_matched_zips <- mort_grouped_zip %>% 
  filter(ZIP_Code %in% air_qual_by_75_percent$zip) 

mort_matched_zips 
```

\newpage

#3. Cleaning data 
Creating new data set with mortality and pollution columns for all zip codes with a critical level of PM2.5 pollution + died from CLD 
```{r }
final_data <- cbind(air_qual_by_75_percent, mort_matched_zips)
final_data <- final_data %>% 
  select( county, zip, avg_PM_per_zip, sum_count)

#cleaning 
final_data$sum_count[is.na(final_data$sum_count)] <- NA
final_data$avg_PM_per_zip <- round(final_data$avg_PM_per_zip, 2)
final_data$zip <-as.character(final_data$zip)
head(final_data)
#final_data[is.na(final_data)] <- 0 

```

\newpage

#4. Data dictionary

```{r}

data_dictionary <- data.frame("Variable name" = c("zip", "avg_PM_per_zip", 
                                                  "sum_count", 
                                                  "critical_level_pollution",
                                                  "county"),
                              "Data type" = c("Character", "Numeric", "Numeric", 
                                              "Numeric", "Character"),
                              "Description" = c("California zip codes", 
                              "Average PM2.5 per zip code", 
                              "Total count of CLDRM for the 2009-2018 period", 
                              "Binary variable to indicate whether or not the zip has a critical value of PM2.5 pollution (at least 11.84)", 
                              "California county names"))

kable(data_dictionary, col.names = c("Variable Name", "Data Type", "Description"), 
      caption = "Data Dictionary for CA CLD mortality per 
      zip code given PM2.5 air pollution rate")

```

\newpage 



#5. One or more tables with descriptive statistics for 4 data elements
```{r}
#table of average pollution & counts CLDRM

myvars <- c("avg_PM_per_zip", "sum_count")

tab_1 <- CreateTableOne(vars = myvars, data = final_data,)

tab_1

```

```{r}
#table of average pollution and deaths per county

avgpm_mort_county<-final_data %>% 
  group_by(county) %>% 
  summarise(avg_pm25=mean(avg_PM_per_zip), death = sum(sum_count, na.rm = T))
avgpm_mort_county <- avgpm_mort_county %>% arrange(desc(avg_pm25))

kable(avgpm_mort_county)

#myvars_2 <- c( "avg_pm25", "death")
#tab_2 <- CreateTableOne(vars = myvars_2, data = avgpm_mort_county, strata = "county")
#tab_2

```

